---
output:
  word_document:
bibliography: MyBib.bib
csl: JAE.csl
header-includes: \usepackage{graphicx}
    \usepackage{setspace}
    \doublespacing
    \usepackage{lineno}
    \linenumbers
---
```{r setup, include = F}

rm(list=ls())

#Load knitr
library(knitr)

#Set some options
options(na.action='na.fail')
opts_knit$set(root.dir = '..')
opts_chunk$set(echo=FALSE,
               warning=FALSE,
               message=FALSE,
               fig.width=8,
               fig.height=8)

```

```{r load and clean, include = F}

#Load and clean data
source("Scripts/SpurginBosse_Hapmap_LOAD.r")
source("Scripts/SpurginBosse_Hapmap_FUNCTIONS.r")
source("Scripts/SpurginBosse_Hapmap_CLEAN.r")
```



#Supplementary material for "A continental-scale analysis of genomic variation in a songbird"


```{r FigS1}
FigS1 <- ggplot(ld2,aes(x = dist/1000,y = meanR2))+
  geom_line()+
  theme_bw()+
  facet_wrap(~pop)+
  xlab("Distance (kb)")+
  ylab(expression("R"^2))
FigS1
```

**Figure S1** Linkage disequilibrium in European great tit populations. Lines are means of $R^2$ values for all pairs of markers within 100bp distance bins.


```{r FigS2,fig.width = 7,fig.height = 12}

for(i in 2:9)
{
  assign(
    paste0("FigS2",LETTERS[i-1]),
    structureplot(get(paste0("admix",i)),pops,i,xlab = F)+
      ylab(paste("K =", i))
        )
}

FigS2I <- structureplot(admix10,pops,10,xlab = T)+
  ylab("K = 10")

multiplot(FigS2A,FigS2B,FigS2C,FigS2D,FigS2E,FigS2F,FigS2G,FigS2H,FigS2I,layout = matrix(c(1:8,9,9),nrow = 10))
```
**Figure S2** Admixture analyses at $K$ = 2 to 10 in European Great tit populations.







```{r FigS3,fig.width = 5,fig.height = 5}
FigS3 <- ggplot(cv,aes(x = K,y = CV))+
  geom_line()+
  theme_bw()+
  xlab(expression(italic("K")))+
  ylab("Cross-validation error")+
  scale_x_continuous(breaks = c(1:10))
FigS3
```

**Figure S3** Cross-validation error from admixture analyses at $K$ = 1 to 10.



```{r Fig S4,fig.width = 5,fig.height = 5}
m = 10
vars <- rep(NA,m+1)
vars[1] <- calcVarExplain("Data/Treemix/HapMap")$VarExplain
for(i in c(2:(m+1)))
{
  vars[i] <- calcVarExplain(paste0("Data/Treemix/HapMap_",i-1))$VarExplain
}

ddS4 <- data.frame(M = c(0:10),vars = vars)
FigS4 <- ggplot(ddS4,aes(x = M,y = vars))+
  geom_line()+
  theme_bw()+
  xlab("Number of migration events")+
  ylab("Proportion relatedness explained")+
  scale_x_continuous(breaks = c(0:10))
FigS4
```

**Figure S4** Proportion of relatedness explained among populations in relation to the number of migration events allowed in Treemix analyses.
