---
output:
  pdf_document:
bibliography: MyBib.bib
csl: PLoSGenetics.csl
header-includes: \usepackage{graphicx}
    \usepackage{setspace}
    \doublespacing
    \usepackage{lineno}
    \linenumbers
    \usepackage{pdflscape}
    \newcommand{\blandscape}{\begin{landscape}}
    \newcommand{\elandscape}{\end{landscape}}

---
```{r setup, include = F}

rm(list=ls())

#Load knitr
library(knitr)

#Set some options
options(na.action='na.fail')
opts_knit$set(root.dir = '..')
opts_chunk$set(echo=FALSE,
               warning=FALSE,
               message=FALSE,
               fig.width=8,
               fig.height=8)

```

```{r load and clean, include = F}

#Load and clean data
source("Scripts/SpurginBosse_Hapmap_LOAD.r")
source("Scripts/SpurginBosse_Hapmap_FUNCTIONS.r")
source("Scripts/SpurginBosse_Hapmap_CLEAN.r")
```



#Supplementary material for "A continental-scale analysis of genomic variation in a songbird"

**Table S1** Sampling details for the `r nrow(ll)` European great tit populations. "Code" corresponds to labels in Figure 1 of the main text.

```{R Table S1}
Table1 <- ll %>%
  mutate(N = table(pops$Pop),
         Pop = str_replace(Pop,"Tartu_Estonia","Kilingi-Nõmme_Estonia")) %>%
  dplyr::rename(Population = Pop,Latitude = Lat,Longitude = Long,Region = Country) %>%
  arrange(Region) %>%
  mutate(Code = c(1:nrow(ll))) %>%
  select(Code,Population,Region,N,Latitude,Longitude)

kable(Table1,digits = 3,row.names = F,format = "latex", booktabs = TRUE) %>%
          kable_styling(latex_options = "scale_down")


```


\pagebreak

**Table S2** Cross-validation error from Admixture analyses at $K$ = 1 to 10. Related to Figure 2.


```{r Table S3}
TableS2 <- cv %>%
  dplyr::rename(`Cross-validation error` = CV)
  
kable(TableS2,digits = 4,row.names = F,format = "latex", booktabs = TRUE)
```

\pagebreak


**Table S3** Proportion of relatedness explained among populations in relation to the number of migration events allowed in TreeMix analyses. Related to Figure 2.

```{r Table S3}
m = 10
vars <- rep(NA,m+1)
vars[1] <- calcVarExplain("Data/Treemix/HapMap")$VarExplain
for(i in c(2:(m+1)))
{
  vars[i] <- calcVarExplain(paste0("Data/Treemix/HapMap_",i-1))$VarExplain
}

TableS3 <- tibble(`Number of migration events` = c(0:10),`Varance explained` = vars)

kable(TableS3,digits = 4,row.names = F,format = "latex", booktabs = TRUE)

```



\pagebreak
\blandscape

**Table S4** Details of outlier $F_{ST}$ regions in European great tit populations. For each region, $F_{ST}$ was calculated for each population versus Turkey. The "number of outliers" represents the number of *populations* in which each region was detected as an outlier. All genes found within 500kb outlier windows are displayed.

```{r Table S4}
# temp <- strsplit(outliers$Window,split = " ")
# TS2 <- cbind(matrix(unlist(temp),ncol = 2,byrow=T),outliers$nhits,outliers$Countries)
# TS2[,2] <- paste(round(as.numeric(TS2[,2])/10e05,1))
# 
# colnames(TS2) <- c("Chromosome","Position (MB)","Number of outliers","Outlier comparisons")
# TS2 <- TS2[order(as.numeric(TS2[,"Number of outliers"]),decreasing = T),]
# TS2[,"Outlier comparisons"] <- gsub("_"," ",TS2[,"Outlier comparisons"])
# TS2[,"Chromosome"] <- gsub("30","1A",TS2[,"Chromosome"])
# TS2[,"Chromosome"] <- gsub("33","4A",TS2[,"Chromosome"])
# 
# write.table(TS2,"Data/TS2.txt",row.names = F,quote = F,sep = "\t")

TS4 <- read_tsv("Data/TS2.txt") %>%
  dplyr::rename("Position (MB)" = Pos,"Number of outliers" = Nhits,"Outlier comparisons" = Outlier_comparisons) %>%
  mutate_all(funs(str_replace(.,"\\."," ")))


kable(TS4,format = "latex", booktabs = TRUE) %>%
          kable_styling(latex_options = "scale_down")

```

\pagebreak
\elandscape


```{r FigS1}

FigS1 <- ld2 %>%
  ggplot(aes(x = dist/1000,y = meanR2))+
  geom_line()+
  facet_wrap(~Pop)+
  xlab("Distance (kb)")+
  ylab(expression("R"^2))
FigS1
```

**Figure S1** Linkage disequilibrium in European great tit populations. Lines are means of $R^2$ values for all pairs of markers within 100bp distance bins. Related to Figure 2.




\pagebreak


```{r FigS2,fig.width = 5,fig.height = 5}
FigS2 <- ggplot(pd,aes(x = dist/1000,y = FST,col = Island))+
            geom_point()+
            theme(legend.position = "none")+
            ylab(expression(italic(F)[ST]))+
            xlab("Geographic distance (km)")+
            scale_colour_manual(values = c("darkblue","grey"))

FigS2
```

**Figure S2** Pairwise $F_{ST}$ in relation to geographic distance among European great tit populations. Pairwise comparisons involving the island populations of Corsica, Sardinia and Crete are coloured in dark blue. Related to Figure 2.


\pagebreak


```{r FigS3,fig.width = 7,fig.height = 9}

for(i in 2:9)
{
  assign(
    paste0("FigS3",LETTERS[i-1]),
    structureplot(get(paste0("admix",i)),pops,i,xlab = F)+
      ylab(paste("K =", i))
        )
}

FigS3I <- structureplot(admix10,pops,10,xlab = T)+
  ylab("K = 10")

plot_grid(FigS3A,FigS3B,FigS3C,FigS3D,FigS3E,FigS3F,FigS3G,FigS3H,FigS3I, align = "v",ncol = 1,rel_heights = c(rep(1,8),2.5))
```

**Figure S3** Admixture analyses at $K$ = 2 to 10 in European great tit populations. Related to Figure 2.

\pagebreak



\pagebreak

```{r FigS4, fig.width = 5, fig.height = 7}
FigS5 <- tibble(Value = unlist(pca[,c(3:6)]),
                  PC = str_c("PC",rep(c(1:4),each = nrow(pca))),
                  Pop = rep(pca$Country,4)) %>%
ggplot(aes(x = Pop, y = Value, col = Pop))+
  geom_jitter(width = 0.2)+
  theme(axis.text.x = element_text(angle = 90,vjust=0.5,hjust=1),
        legend.position = "none")+
  facet_wrap(~PC,nrow = 4)+
  xlab("")+
  ylab("Score")

FigS5
```

**Figure S4** Results from PCA analysis of European great tit populations. Related to Figure 2.






\pagebreak

```{r Figure S5,fig.width=8,fig.height=4}
FigS7A <- dw %>%
  group_by(Window) %>%
  dplyr::summarise(piTurkey = mean(pi_Turkey),
              pi = mean(pi_pop1)) %>%
  ggplot(aes(x = piTurkey,y = pi))+
                 geom_point(col = "grey")+
  geom_point(data = outlierhaps,aes(col = nhf, x = pi))+
  theme(legend.position = "none")+
  scale_color_manual(values = c("red","navy"))+
  xlab(expression(pi["Turkey"]))+
  ylab(expression(pi["All populations"]))+
  xlim(c(0.15,0.5))+
  ylim(c(0.15,0.5))+
  ggtitle("A")


FigS7B <- ggplot(outlierhaps,
               aes(x = log10(Markers),y = log10(Richness),col = nhf,size = N_hits))+
                 geom_point()+
  theme(legend.position = "none")+
  scale_color_manual(values = c("red","navy"))+
  xlab("Log marker density")+
  ylab("Log number of haplotypes")+
  ggtitle("B")

multiplot(FigS7A,FigS7B,layout = matrix(c(1:2),nrow = 1))

```

**Figure S5** Genetic diversity in $F_{ST}$ outlier regions in European great tit populations. Each point is an outlier window, with point size scaled by the number of populations the outlier was found in - "shared" haplotypes are coloured red and "unique" haplotypes dark blue (see main text for details). In **A**, Nucleotide diversity for each window is calculated for the proposed refugial population in Turkey, and plotted against the mean for all other populations. In **B**, SNP marker density is plotted against haplotype richness for each region. Related to Figure 4.



\pagebreak

```{r FigS6}
knitr::include_graphics("../Figures/Haplotype_structure_figure.pdf")
```

**Figure S6** Haplotype structure underlying regions of high differentiation.
**A** IBS clustering of all European haplotypes in the region containing a signal of high differentiation with Turkey specific to the Oulu population. The red arrow indicates the haplotype at highest frequency in Finland. Colour indicates IBS similarity (range 0: white - 1: dark blue). **B** IBS clustering of all European haplotypes in the region on chromosome 2 containing a signal of high differentiation with Turkey in multiple populations. **C** Frequency of all identified haplotypes within each population. The red arrow indicates the high-frequency haplotype within Finland. **D** Frequency of all identified haplotypes spanning the region on chromosome 2 within each population. Related to Figure 5; Figure S4.
