---
output:
  pdf_document:
bibliography: MyBib.bib
csl: PLoSGenetics.csl
header-includes: \usepackage{graphicx}
    \usepackage{setspace}
    \doublespacing
    \usepackage{lineno}
    \linenumbers
    \usepackage{pdflscape}
    \newcommand{\blandscape}{\begin{landscape}}
    \newcommand{\elandscape}{\end{landscape}}

---
```{r setup, include = F}

rm(list=ls())

#Load knitr
library(knitr)

#Set some options
options(na.action='na.fail')
opts_knit$set(root.dir = '..')
opts_chunk$set(echo=FALSE,
               warning=FALSE,
               message=FALSE,
               fig.width=8,
               fig.height=8)

```

```{r load and clean, include = F}

#Load and clean data
source("Scripts/SpurginBosse_Hapmap_LOAD.r")
source("Scripts/SpurginBosse_Hapmap_FUNCTIONS.r")
source("Scripts/SpurginBosse_Hapmap_CLEAN.r")
```



#Supplementary material for "The great tit HapMap project: a continental-scale analysis of genomic variation in a songbird"

**Table S1** Sampling details for the `r nrow(ll)` European great tit populations.

```{R Table S1}
Table1 <- ll
Table1$LD <- NULL
colnames(Table1) <- c("Population","Latitude","Longitude")


temppops <- as.character(paste(Table1$Population))
countries <- as.character(paste(Table1$Pop))


temppops[temppops %in% c("Antwerp_Belgium")] <- "Antwerp"
temppops[temppops %in% c("Cambridge_UK")] <- "Cambridge"
temppops[temppops %in% c("Wytham_UK")] <- "Oxford"
temppops[temppops %in% c("Mariola_Spain")] <- "Mariola"
temppops[temppops %in% c("Font_Roja_Spain")] <- "Font Roja"
temppops[temppops %in% c("Gotland_Sweden")] <- "Gotland"
temppops[temppops %in% c("Harjavalta_Finland")] <- "Harjavalta"
temppops[temppops %in% c("Oulu_Finland")] <- "Oulu"
temppops[temppops %in% c("Montpellier_France")] <- "Montpellier"
temppops[temppops %in% c("La_Rouviere_France")] <- "La Rouviere"
temppops[temppops %in% c("Loch_Lomond_Scotland")] <- "Loch Lomond"
temppops[temppops %in% c("Pilis_Mountains_Hungary")] <- "Pilis Mountains"
temppops[temppops %in% c("Pirio_Muro_Corsica")] <- "Pirio Muro"
temppops[temppops %in% c("Radolfzell_Germany")] <- "Radolfzell"
temppops[temppops %in% c("Seewisen_Germany")] <- "Seewisen"
temppops[temppops %in% c("Roekelse_Bos")] <- "Roekelse Bos"
temppops[temppops %in% c("Vlieland_NL")] <- "Vlieland"
temppops[temppops %in% c("Tartu_Estonia")] <- "Tartu"
temppops[temppops %in% c("Velky_Kosir_Czech_Republic")] <- "Velky Kosir"
temppops[temppops %in% c("Vienna_Austria")] <- "Vienna"
temppops[temppops %in% c("Zurich_Switzerland" )] <- "Zurich"                             
temppops[temppops %in% c("Zvenigorod_Russia")] <- "Zvenigorod"

countries[countries %in% c("Antwerp_Belgium")] <- "Belgium"
countries[countries %in% c("Cambridge_UK","Wytham_UK")] <- "England"
countries[countries %in% c("Font_Roja_Spain","Mariola_Spain")] <- "Spain"
countries[countries %in% c("Gotland_Sweden")] <- "Sweden"
countries[countries %in% c("Harjavalta_Finland","Oulu_Finland")] <- "Finland"
countries[countries %in% c("La_Rouviere_France" ,"Montpellier_France")] <- "France"
countries[countries %in% c("Loch_Lomond_Scotland")] <- "Scotland"
countries[countries %in% c("Pilis_Mountains_Hungary")] <- "Hungary"
countries[countries %in% c("Pirio_Muro_Corsica")] <- "Corsica"
countries[countries %in% c("Radolfzell_Germany","Seewisen_Germany")] <- "Germany"
countries[countries %in% c("Roekelse_Bos","Westerheide")] <- "Netherlands"
countries[countries %in% c("Vlieland_NL")] <- "Netherlands (Vlieland)"
countries[countries %in% c("Tartu_Estonia")] <- "Estonia"
countries[countries %in% c("Velky_Kosir_Czech_Republic")] <- "Czech Republic"
countries[countries %in% c("Vienna_Austria")] <- "Austria"
countries[countries %in% c("Zurich_Switzerland" )] <- "Switzerland"                             
countries[countries %in% c("Zvenigorod_Russia")] <- "Russia"
countries[countries %in% c("Romania", "Bulgaria")] <- "Balkans"      

Table1$Population <- temppops  
Table1$Region <- factor(countries,
                  levels = c(
                    "Scotland",
                    "England",
                    "Spain",
                    "France",
                    "Belgium",
                    "Netherlands",
                    "Netherlands (Vlieland)",
                    "Switzerland",
                    "Germany",
                    "Corsica",
                    "Sardinia",
                    "Italy",
                    "Austria",
                    "Czech Republic",
                    "Hungary",
                    "Sweden",
                    "Finland",
                    "Estonia",
                    "Crete",
                    "Balkans",
                    "Turkey",
                    "Russia"))

Table1$N <- table(pops$o_pop)

Table1 <- Table1[order(Table1$Region),c("Population","Region","N","Latitude","Longitude")]
kable(Table1,digits = 3,row.names = F,format = "latex", booktabs = TRUE) %>%
          kable_styling(latex_options = "scale_down")


```


\pagebreak
\blandscape

**Table S2** Details of outlier $F_{ST}$ regions in European great tit populations. For each region, $F_{ST}$ was calculated for each population versus Turkey. The "number of outliers represents" the number of *populations* in which each region was detected as an outlier. All genes found within 500kb outlier windows are displayed.

```{r Table S2}
# temp <- strsplit(outliers$Window,split = " ")
# TS2 <- cbind(matrix(unlist(temp),ncol = 2,byrow=T),outliers$nhits,outliers$Countries)
# TS2[,2] <- paste(round(as.numeric(TS2[,2])/10e05,1))
# 
# colnames(TS2) <- c("Chromosome","Position (MB)","Number of outliers","Outlier comparisons")
# TS2 <- TS2[order(as.numeric(TS2[,"Number of outliers"]),decreasing = T),]
# TS2[,"Outlier comparisons"] <- gsub("_"," ",TS2[,"Outlier comparisons"])
# TS2[,"Chromosome"] <- gsub("30","1A",TS2[,"Chromosome"])
# TS2[,"Chromosome"] <- gsub("33","4A",TS2[,"Chromosome"])
# 
# write.table(TS2,"Data/TS2.txt",row.names = F,quote = F,sep = "\t")

TS2 <- read.table("Data/TS2.txt", header = T,as.is = T)
colnames(TS2) <- c("Chromosome","Position (MB)","Number of outliers","Outlier comparisons","Genes")
TS2 <- as.matrix(TS2)
TS2 <- data.frame(gsub("."," ",TS2,fixed = T))


kable(TS2,format = "latex", booktabs = TRUE) %>%
          kable_styling(latex_options = "scale_down")

```

\pagebreak
\elandscape



```{r FigS1}
FigS1 <- ggplot(ld2,aes(x = dist/1000,y = meanR2))+
  geom_line()+
  theme_bw()+
  facet_wrap(~pop)+
  xlab("Distance (kb)")+
  ylab(expression("R"^2))
FigS1
```

**Figure S1** Linkage disequilibrium in European great tit populations. Lines are means of $R^2$ values for all pairs of markers within 100bp distance bins.




\pagebreak


```{r FigS2,fig.width = 5,fig.height = 5}
FigS2 <- ggplot(pd,aes(x = dist/1000,y = FST,col = Island))+
            geom_point()+
            theme_bw()+
            theme(legend.position = "none")+
            ylab(expression(italic(F)[ST]))+
            xlab("Geographic distance (km)")+
            scale_colour_manual(values = c("darkblue","grey"))

FigS2
```

**Figure S2** Pairwise $F_{ST}$ in relation to geographic distance among European great tit populations. Pairwise comparisons involving the island populations of Corsica, Sardinia and Crete are coloured in dark blue.


\pagebreak


```{r FigS3,fig.width = 7,fig.height = 10}

for(i in 2:9)
{
  assign(
    paste0("FigS3",LETTERS[i-1]),
    structureplot(get(paste0("admix",i)),pops,i,xlab = F)+
      ylab(paste("K =", i))
        )
}

FigS3I <- structureplot(admix10,pops,10,xlab = T)+
  ylab("K = 10")

multiplot(FigS3A,FigS3B,FigS3C,FigS3D,FigS3E,FigS3F,FigS3G,FigS3H,FigS3I,layout = matrix(c(1:8,9,9),nrow = 10))
```

**Figure S3** Admixture analyses at $K$ = 2 to 10 in European Great tit populations.

\pagebreak






```{r FigS4,fig.width = 5,fig.height = 5}
FigS4 <- ggplot(cv,aes(x = K,y = CV))+
  geom_line()+
  theme_bw()+
  xlab(expression(italic("K")))+
  ylab("Cross-validation error")+
  scale_x_continuous(breaks = c(1:10))
FigS4
```

**Figure S4** Cross-validation error from admixture analyses at $K$ = 1 to 10.


\pagebreak

```{r FigS5, fig.width = 5, fig.height = 7}
FigS5 <- data.frame(Value = unlist(pca[,c(3:6)]),
                  PC = paste0("PC",rep(c(1:4),each = nrow(pca))),
                  Pop = rep(pca$V1,4)) %>%
ggplot(aes(x = Pop, y = Value, col = Pop))+
  geom_jitter(width = 0.2)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90),
        legend.position = "none")+
  facet_wrap(~PC,nrow = 4)+
  xlab("")+
  ylab("")

FigS5
```

**Figure S5** Results from PCA analysis of European great tit populations.



\pagebreak

```{r Fig S6,fig.width = 5,fig.height = 5}
m = 10
vars <- rep(NA,m+1)
vars[1] <- calcVarExplain("Data/Treemix/HapMap")$VarExplain
for(i in c(2:(m+1)))
{
  vars[i] <- calcVarExplain(paste0("Data/Treemix/HapMap_",i-1))$VarExplain
}

ddS6 <- data.frame(M = c(0:10),vars = vars)
FigS6 <- ggplot(ddS6,aes(x = M,y = vars))+
  geom_line()+
  theme_bw()+
  xlab("Number of migration events")+
  ylab("Proportion relatedness explained")+
  scale_x_continuous(breaks = c(0:10))
FigS6
```

**Figure S6** Proportion of relatedness explained among populations in relation to the number of migration events allowed in Treemix analyses.


\pagebreak






```{r Figure S6,fig.width = 8,fig.height = 6}

dw$Window <- paste(dw$scaffold,dw$start)
temp <- subset(subset(dw,!(scaffold %in% c(36))),pop1!="Balkans")
temp$FST[temp$FST < 0] <- 0
temp$pop1 <- factor(temp$pop1,levels = names(tapply(temp$FST,temp$pop1,mean)[order(tapply(temp$FST,temp$pop1,mean))]))

temp2 <- subset(temp,zFST > 10) %>%
  ddply(.(Window),
        summarise,
        n_hits = length(zFST))

x1 <- droplevels(subset(subset(dw,zFST > 10),Window %in% temp2$Window))
x1$nhits <- tapply(x1$zFST,x1$Window,length)[x1$Window]
x1$nhf <- ifelse(x1$nhits > 2,"Shared","Unique")

FigS7<- ggplot(temp,aes(y = zFST,x = order,col = factor(scaffold)))+
  theme_bw()+
  theme(axis.text.x = element_blank(),
        legend.position = "none")+
    geom_line()+
    geom_vline(data = x1,aes(xintercept = order,lty = nhf),col = "red",alpha = 0.15,lwd = 1.2)+
  facet_wrap(~pop1)+
              ylab(expression(italic(zF)[ST]))+
            xlab("Genomic position")+
  scale_color_manual(values = rep(c("navy","grey"),30))
FigS7
```

**Figure S7** Landscapes of relative genomic differentiation in European great tit populations. $zF_{ST}$ across the genome is averaged in 500kb windows, with each panel displaying a pairwise comparison between with the proposed refugial population in Turkey. Red lines represent $F_{ST}$ outliers (windows with mean $zF_{ST}$ > 10) shared across more than two comparisons (solid red lines), or unique to one or two comparisons (dashed red lines).


\pagebreak

```{r Figure S8,fig.width=8,fig.height=4}
temp <- ddply(dw,
              .(Window),
              summarise,
              piTurkey = mean(pi_Turkey),
              pi = mean(pi_pop1))


FigS9A <- ggplot(temp,
               aes(x = piTurkey,y = pi))+
                 geom_point(col = "grey")+
  geom_point(data = outlierhaps,aes(col = nhf))+
    theme_bw()+
  theme(legend.position = "none")+
  scale_color_manual(values = c("red","navy"))+
  xlab(expression(pi["Turkey"]))+
  ylab(expression(pi["All populations"]))+
  xlim(c(0.15,0.5))+
  ylim(c(0.15,0.5))+
  ggtitle("A")


FigS9B <- ggplot(outlierhaps,
               aes(x = log10(Markers),y = log10(Richness),col = nhf,size = N_hits))+
                 geom_point()+
    theme_bw()+
  theme(legend.position = "none")+
  scale_color_manual(values = c("red","navy"))+
  xlab("Log marker density")+
  ylab("Log number of haplotypes")+
  ggtitle("B")

multiplot(FigS9A,FigS9B,layout = matrix(c(1:2),nrow = 1))

```

**Figure S8** Genetic diversity in $F_{ST}$ outlier regions in European great tit populations. Each point is an outlier window, with point size scaled by the number of populations the outlier was found in - "shared" haplotypes are coloured red and "unique" haplotypes dark blue (see main text for details). In **A**, Nucleotide diversity for each window is calculated for the proposed refugial population in Turkey, and plotted against the mean for all other populations. In **B**, SNP marker density is plotted against haplotype richness for each region.



\pagebreak

```{r FigS8}
knitr::include_graphics("../Haplotype_structure_figure.pdf")
```

**Figure S8** Haplotype structure underlying regions of high differentiation.
**A** IBS clustering of all European haplotypes in the region containing a signal of high differentiation with Turkey specific to the Oulu population. The red arrow indicates the haplotype at highest frequency in Oulu. Colour indicates IBS similarity (range 0:white - 1: dark blue). **B** IBS clustering of all European haplotypes in the region on chromosome 2 containing a signal of high differentiation with Turkey in multiple populations. **C** Frequency of all identified haplotypes within each population. The red arrow indicates the high-frequency haplotype within Oulu. **D** Frequency of all identified haplotypes spanning the region on chromosome 2 within each population.
