---
output:
  word_document:
bibliography: MyBib.bib
csl: JAE.csl
header-includes: \usepackage{graphicx}
    \usepackage{setspace}
    \doublespacing
    \usepackage{lineno}
    \linenumbers
---
```{r setup, include = F}

rm(list=ls())

#Load knitr
library(knitr)

#Set some options
options(na.action='na.fail')
opts_knit$set(root.dir = '..')
opts_chunk$set(echo=FALSE,
               warning=FALSE,
               message=FALSE,
               fig.width=8,
               fig.height=8)

set.seed(200285)
```



```{r load and clean, include = F}

#Load and clean data
source("Scripts/SpurginBosse_Hapmap_LOAD.r")
source("Scripts/SpurginBosse_Hapmap_FUNCTIONS.r")
source("Scripts/SpurginBosse_Hapmap_CLEAN.r")
```


#A continental-scale analysis of genomic variation in a songbird

Lewis G. Spurgin^1,2^\*, Mirte Bosse^3,4^\*, Veronika N. Laine^3^, The Great Tit HapMap Consortium, Martien A.M. Groenen^4^, Kees van Oers^3^, Ben C. Sheldon^1^, Marcel E.  Visser^3^ and Jon Slate^5^\**


1 School of Biological Sciences, University of East Anglia, Norwich Research Park, United Kingdom

2 Edward Grey Institute, University of Oxford, United Kingdom

3 Department of Animal Ecology, Netherlands Institute of Ecology (NIOO-KNAW), Wageningen, the Netherlands

4 Animal Breeding and Genomics Centre, Wageningen University, the Netherlands

5 Department of Animal and Plant Sciences, University of Sheffield, United Kingdom

\* These authors contributed equally to this manuscript

\** **Correspondence**: j.slate@sheffield.ac.uk

\pagebreak




##Abstract

A major aim of evolutionary biology is to understand why patterns of genetic variation difffer within genomes, and among populations and species. Widespread continental species are interesting models for studying how evolution shapes patterns of genomic divergence, as they are associated with recent and rapid colonisations, large population sizes and high levels of gene flow. We studied patterns of genomic variation among `r nrow(ll)` great tit (*Parus major*) in Europe. We found that genome-wide variation was consistent with a recent colonisation across Europe from a single refugium in the Balkans and/or Turkey. The highest levels of genome-wide structure were found between island populations in Crete, Sardinia and Corsica, most likely as a resut of mild bottlenecks and reduced gene flow among these populations. We then found that the landscape of genomic differentiation was highly heterogeneous, with clear "islands of divergence" among both closely related and structured great tit populations. 

##Introduction

Since the first studies of allozyme variation in humans [@Harris1966] and Drosophila [@Lewontin1966; @Hubby1966], there has been great interest in explaining how evolutionary and ecological processes shape the patterns of genetic variation observed within and among natural populations. One focus of research and debate in this area has been on quantifying the roles of adaptive and neutral processes in explaining observed levels of genetic variation. However, adaptation does not occur in isolation, but against a background of recombination, gene flow, and genetic drift. More recently there has been increased effort in understanding how these fundamental evolutionary forces operate in concert to generate and maintain the levels of genetic diversity commonly observed in natural populations.

The arrival of high-throughput sequencing and subsequent characterisation of genome-wide variation across large numbers of individuals has revealed that at the genetic level, patterns of variation and divergence among natural populations and species are highly heterogeneous. A feature of these "genomic landscapes" of differentiaiton that has received particular attention is the presence of so-called "islands of divergence"; outlier regions of the genome with high levels of divergence estimated from statistics such as $F_{ST}$ or $d_{xy}$ [e.g. @Turner2005; @Nadeau2011; @Renaut2013; @Poelstra2014]. Initially these regions were termed "islands of speciation", and were thought to arise as a result of reduced gene flow in genomic regions associated with reproductive isolation [@Turner2005]. Subsequent modelling and analysis revealed that highly heterogeneous patterns of genomic divergence in the complete absence of gene flow, as a result of linked selection on beneficial or deleterious mutations [@Noor2009; @Cruickshank2014].

Comparing patterns of genomic differentiation among sets of populations and/or species at different stages of the speciation continuum is a powerful way of disentangling the forces that shape variation among populations. Martin et al. [-@Martin2013] showed that, in *Heliconius* butterflies, 

Perhaps the most striking results from comparisons of genomic variation among populations and species have come from recent research in birds, which has revealed that differentiation landscapes are conserved across populations, species and even across avian families [@Burri2015; @Vijay2017; @VanDoren2017]. This pattern has arisen, at least in part, as a result of a highly conserved recombination landscape in birds.

The European great tit (*Parus major major*) is an excellent model for ecological and evolutionary study. Phyleogeographic research using mitochondrial and microsatellite loci suggests that this species has has recently expanded across its European range, possible from a single refugium in the Balkans. Contemporary populations are characterised by large effective population sizes and high levels of gene flow among populations. The genome of the great tit has recently been sequences, and a high density (650K) panel of SNP markers has been developed. A recent study of two European populations using this marker panel suggests that rapid adaptation has occurred at the genomic and phenotypic levels [@Bosse2017]. 

In this study we examined how selection, gene flow and population history have shaped patterns of genomic variation across recently colonised European great tit populations. Using a large SNP panel typed across `r nrow(ll)` European populations, we first characterise variation within and among populations, in order to infer patterns of population history. We then examine how variation is partitioned across the genome, and test the hypothesis that highly divergent genomic regions have arisen as a result of variation in recombination rate across the genome. Finally, we examine how the relationship between genomic divergence and recombination varies according to geography, with the aim of inferring how gene flow and drift interact with selection to shape genomic landscapes in natural populations.

##Methods

####Sampling and molecular methods
Samples were collected from `r nrow(ll)` populations from `r length(unique(pops$V1))` countries across Europe (Fig. 1A; Table 1). Birds were trapped from nest boxes, or using mist nets, and ringed with a uniquely numbered aluminium ring. Blood was taken via brachial venipuncture, and stored in either 1 ml Cell Lysis Solution (Gentra Puregene Kit, Qiagen, USA) or Queen’s buffer. All samples were genotyped using a custom-made Affymetrix® great tit 650K SNP chip at Edinburgh Genomics (Edinburgh, United Kingdom). After filtering, a total of `r nrow(pops)` samples typed at `r nrow(markers)` SNPs were retained for analysis (Table 1).


####Analyses
Unless stated otherwise, all population genetic statistics were calculated in PLINK version 1.9 [@Purcell2007], and downstream analysis and plotting was carried out in R version 3.3 [@RDevelopmentCoreTeam2011]. In each population, we estimated linkage disequilibrium ($R^2$) for each pair of markers within 50kb on the same chromosome, and compared this to physical distance between marker pairs. For a simple comparison of short-range LD among populations, we also calculated, for each population, mean LD between all pairs of markers within 1kb. We then calculated observed heterozygosity for each SNP and population using a reduced SNP dataset, which was pruned based on linkage disequilibrium to remove all markers with $R^2$ > 0.1, then thinned with a probability of retaining each variant of 0.25.

We calculated mean $F_{ST}$ for all loci between each pair of populations using the pruned and thinned dataset described above. Pairwise $F_{ST}$ was compared to geographic distance between populations using Mantel tests, implemented in the Ecodist package in R [@Goslee2007]. We also estimated population structure using Admixture version 1.3, with default settings [@Alexander2009]. We varied values of $K$ from one to ten; by which time increasing values of $K$ provided no informative information about population structure (see results). Model support for each value of $K$ was estimated by calculating 5-fold cross-validation error. Finally, we visualised the evolutionary history among European great populations by generating a maximum likelihood tree in TreeMix version 1.13 [@Pickrell2012]. We rooted the tree using a sample of *P. minor* individuals sampled from Amur, Russia. We fitted models allowing for range of migration events (0-10), and used a window size of 500 SNPs [@Pickrell2012]. To assess model fit, we calculated the proportion of variance relatedness between populations explained by each model, using the equation 30 in Pickrell *et al.* [-@Pickrell2012].

We examined the genomic landscape of differentiation across European great tit populations by calculating marker-based $F_{ST}$ in Plink. For plotting and analysis, we calculted mean $F_{ST}$ values in 200kb bins, using custom R and python scripts. We first calculated windowed $F_{ST}$ between the most structured populations, as defined by Admixture analysis (hereafter "marginal populations"), and between populations within a panmictic cluster in central Europe (hereafter "central European populations"). We tested whether the same regions of the genome were highly structured among these different sets of populations by comaparing windowed $F_{ST}$ values, using Pearson correlations.

Numerous studies have shown that the genomic landscape of diffentiation is related to recombination rate variation (REFS). We tested whether this was the case using a recombination map preivously developed for the great tit using a 10K SNP chip [@vanOers2014]. We imputed recombination rates for all SNPs in our dataset based on genetic distance to markers with known recombination rates, and from this calculated mean recombination rates in 200kb non-overlapping windows. We then compared 200kb averaged recombination rates with 20kkb averaged $F_{ST}$ values, using linear regression. We also specifically tested whether $F_{ST}$ outliers were more likely to occur in regions of low recombination; we defined outlier regions as 200kb bins with mean $F_{ST}$ values greater than the 95th percentile, and tested whether outlier presence/absence was related to recombination rate, using logistic regression.

Finally, we explored how the relationship between genomic differentiation and recombination rate variation was related to geography. We calculated genome-wide differentiation in 200kb bins for all `r nrow(pd)` population pairs, and identified $F_{ST}$ outlier regions as above. Then for every pair of populations we performed linear and logistic regressions of$F_{ST}$ and outlier presence/absence, respectively, against recombination rate, and extracted the slope of these models. We tested whether variation in these slopes were related to pairwise geographic distance between populations, using a Mantel test. We also tested whether the relationships between divergence and recombination were greater in pairwise comparisons involving island populations, were gene flow is assumed to be reduced, using a randomisation test.





##Results


#### Genetic diversity and population history

Sampling locations and sample sizes for each population are given in Table 1. For ease of presentation, we grouped populations into "regions" for plots and analyses (see Table 1 for details). Levels of genetic diversity were generally high and LD declined rapidly in all populations, but we observed substantial differences in genetic diversity among populations. Highest levels of LD were observed in the island populations of Crete (Greece), Sardinia (Italy), Corsica (France) and Vlieland (Netherlands) with lowest levels of LD in central Europe, Turkey and Spain (Fig. 1A; Fig. S1). Similarly, heterozygosity was highest in central Europe, and lowest in Crete, Sardinia and the Balkans (Fig. 1B).

Considering all European great tit populations, there was no significant pattern of isolation-by-distance (Mantel test; r = `r round2(mantel(FST~dist,data = pd)["mantelr"])`, p = `r round2(mantel(FST~dist,data = pd)["pval1"])`). Instead, the highest levels of $F_{ST}$ were found in comparisons involving island the populations of Corsica, Sardinia and Crete (Fig. 2A). Considering only comparisons involving mainland populations, pairwise $F_{ST}$ was still not related to related to geographic distance (r = `r round2(mantel(FST~dist,data = subset(pd,Island == "Mainland"))["mantelr"])`, p = `r round2(mantel(FST~dist,data = subset(pd,Island == "Mainland"))["pval1"])`). We then examined genetic structure in relation to distance from potential refugial populations (only including mainland populations), and found that $F_{ST}$ was significantly related to distance from Turkey (r = `r round2(with(subset(subset(pd,p1 == "Turkey" | p2 == "Turkey"),Island == "Mainland"),cor.test(FST,dist))$estimate)`, p `r round3(with(subset(subset(pd,p1 == "Turkey" | p2 == "Turkey"),Island == "Mainland"),cor.test(FST,dist))$p.value)`; Fig. 2B), but not to distance from Spain (r = `r round2(with(subset(subset(pd,p1 == "Mariola_Spain" | p2 == "Mariola_Spain"),Island == "Mainland"),cor.test(FST,dist))$estimate)`, p = `r round2(with(subset(subset(pd,p1 == "Mariola_Spain" | p2 == "Mariola_Spain"),Island == "Mainland"),cor.test(FST,dist))$p.value)`), or from Italy (r = `r round2(with(subset(subset(pd,p1 == "Italy" | p2 == "Italy"),Island == "Mainland"),cor.test(FST,dist))$estimate)`, p = `r round2(with(subset(subset(pd,p1 == "Italy" | p2 == "Italy"),Island == "Mainland"),cor.test(FST,dist))$p.value)`).

Admixture analysis revealed patterns of genetic structure corresponding to geographical variation in European great tits (Fig. S2). The $K$ = 2 analysis assigned individuals in Sardinia and Corsica to one genetic cluster, and the remaining populations to the second. At $K$ = 3 (the model that best fitted the great tit data; Fig. S3), Spain was separated from the rest of mainland Europe. Increasing values of $K$ resulted of the separation of populations in Scotland ($K$ = 4), Sardinia (from Corsica; $K$ = 5), France ($K$ = 6), Crete ($K$ = 7) and England ($K$ = 8). The admixture output at $K$ = 8 is displayed in Fig. 2C; further increases in $K$ did not generate patterns of structure that corresponded to geographical variation (Fig. S2), and were increasingly less well supported (Fig. S3).

Maximum likelihood analyses implemented in Treemix showed that a model with no migration explained 97.8% of variance in relatedness between populations; increasing the number of migration events substantially improved the percentage of relatedness explained, up to 99.7% when 10 migration events were fitted (Fig. S4). In Figure 3 we display maximum likelihood tree with 2 migration events, after which the variance in relatedness explained plateaued when more migration events were added (Fig. S4). The tree was generally characterised by short branch lengths, with the exception of the island populations of Sardinia and Crete, which were grouped with the population from mainland Italy.  The Spanish and Corsican populations were also grouped together, and were populations from Turkey and the Balkans, and populations from England and Scotland (Fig. 3). The two fitted migration edges both involved Sardinia, involving migration from eastern Europe, and to Corsica (Fig. 3).

#### Genomic landscapes of differentiation
We found that $F_{ST}$ varied markedly across the genome, regardless of whether we considered marginal populations across which structure was higher or across central European populations with very low structure (Fig. 4A). $F_{ST}$ outliers were likely to be found in the same regions across the marginal and central populations (R = `r round2(with(recomb,cor(outlier_cen,outlier_admix)))` $P$ `r round3(with(recomb,cor.test(outlier_cen,outlier_admix))$p.value)`). In addition, $F_{ST}$ outliers were more likely to be found in regions of low recombination across the marginal (GLM; $P$ = `r round3(summary(glm(outlier_admix~log(MEAN_cM),data = recomb,family = "binomial"))$coefficients["log(MEAN_cM)","Pr(>|z|)"])`), but not central ($P$ = `r round(summary(glm(outlier_cen~log(MEAN_cM),data = recomb,family = "binomial"))$coefficients["log(MEAN_cM)","Pr(>|z|)"])`), populations. We found similar results when look at at mean windowed $F_{ST}$ values, rather than presence/absence of outliers. Mean windowed $F_{ST}$ was correlated across these two sets of populations (R = `r round2(with(recomb,cor(fst_cen,fst_admix)))` $P$ `r round3(with(recomb,cor.test(fst_cen,fst_admix))$p.value)`). $F_{ST}$ was negatively related to variation in recombination rate across the marginal populations (R^2^ = `r round2(summary(lm(fst_admix~log(MEAN_cM),data = recomb))$r.squared)`, $P$ `r round3(summary(lm(fst_admix~log(MEAN_cM),data = recomb))$coefficients["log(MEAN_cM)","Pr(>|t|)"])`; Fig. 4C), but no such relationship was found across the central populations ($R^2$ = `r round3(summary(lm(fst_cen~log(MEAN_cM),data = recomb))$r.squared)`, $P$ = `r round2(summary(lm(fst_cen~log(MEAN_cM),data = recomb))$coefficients["log(MEAN_cM)","Pr(>|t|)"])`; Fig. 4D). 

We next quantified how the landscape of genomic differentiation was related to recombination rate for all `r nrow(pd)` pairs of populations. The slope of the relationship between recombination rate was negative in the majority (`r round((sum(pd$fst_outlier < 0)/nrow(pd))*100,0)`%)  $F_{ST}$ outliers were most likely to be found in regions of low recombination in pairwise comparisons involving the island populations of Corsica, Sardinia and Crete, while outliers were found in regions with higher recombination rates in pairwise comparisons involving central European populations (Fig. 5A). Indeed, pairwise comparisons involving islands ($P$ `r round3(wilcox.test(outlier_slope~Island,data=pd)$p.value)`; Fig. S5B). However, the relationship between $F_{ST}$ outliers and recombination rate did not vary according to the geographical distance between populations (r = `r round2(mantel(outlier_slope~dist,data = pd)["mantelr"])`, p = `r round2(mantel(outlier_slope~dist,data = pd)["pval3"])`). 

We found substantial variation in the slope of the relationship between $F_{ST}$ and recombination rate (Fig. S5A), although the slope was negative in the majority (`r round((sum(pd$fst_slope < 0)/nrow(pd))*100,0)`%) of cases. Importantly, the relationship between $F_{ST}$ and recombination rate was non-random with respect to population ID. Nine of the 10 most negative slopes were from pairwise comparisons involving Sardinia, with comparisons involving Corsica and Crete also exhibiting more negative relaitonships between $F_{ST}$ and recombination rate. Accordingly, pairwise comparisons involving island populations had significantly lower slopes than comparisons involving only mainland populations ($P$ `r round3(wilcox.test(fst_slope~Island,data=pd)$p.value)`; Fig. 5B). The relationship between $F_{ST}$ and recombination rate did not vary according to the geographical distance between populations (r = `r round2(mantel(fst_slope~dist,data = pd)["mantelr"])`, p = `r round2(mantel(fst_slope~dist,data = pd)["pval3"])`).

Finally, we tested how the relationship between $F_{ST}$ and recombination rate scaled with genome-wide variaiton in $F_{ST}$ among pairs of populations. We found a strong negative relationship between the $F_{ST}$/recombination slope and genome-wide $F_{ST}$ (r = `r round2(mantel(outlier_slope~FST,data = pd)["mantelr"])`, p < `r round3(mantel(outlier_slope~FST,data = pd)["pval3"])`; Fig. 5C).

##Discussion

##References

<div id="refs"></div>


\pagebreak

##Tables and Figures

**Table 1** Sampling details for the `r nrow(ll)` European great tit populations.

```{R Table 1}
Table1 <- ll
Table1$LD <- NULL
colnames(Table1) <- c("Population","Latitude","Longitude")

countries <- as.character(paste(Table1$Pop))

countries[countries %in% c("Antwerp_Belgium")] <- "Belgium"
countries[countries %in% c("Cambridge_UK","Wytham_UK")] <- "England"
countries[countries %in% c("Font_Roja_Spain","Mariola_Spain")] <- "Spain"
countries[countries %in% c("Gotland_Sweden")] <- "Sweden"
countries[countries %in% c("Harjavalta_Finland","Oulu_Finland")] <- "Finland"
countries[countries %in% c("La_Rouviere_France" ,"Montpellier_France")] <- "France"
countries[countries %in% c("Loch_Lomond_Scotland")] <- "Scotland"
countries[countries %in% c("Pilis_Mountains_Hungary")] <- "Hungary"
countries[countries %in% c("Pirio_Muro_Corsica")] <- "Corsica"
countries[countries %in% c("Radolfzell_Germany","Seewisen_Germany")] <- "Germany"
countries[countries %in% c("Roekelse_Bos","Westerheide_Netherlands")] <- "Netherlands"
countries[countries %in% c("Vlieland_Netherlands")] <- "Netherlands (Vlieland)"
countries[countries %in% c("Tartu_Estonia")] <- "Estonia"
countries[countries %in% c("Velky_Kosir_Czech_Republic")] <- "Czech Republic"
countries[countries %in% c("Vienna_Austria")] <- "Austria"
countries[countries %in% c("Zurich_Switzerland" )] <- "Switzerland"                                                               
countries[countries %in% c("Zvenigorod_Russia")] <- "Russia"                                                               
countries[countries %in% c("Romania", "Bulgaria")] <- "Balkans"      


Table1$Region <- factor(countries,
                  levels = c(
                    "Scotland",
                    "England",
                    "Spain",
                    "France",
                    "Belgium",
                    "Netherlands",
                    "Netherlands (Vlieland)",
                    "Switzerland",
                    "Germany",
                    "Corsica",
                    "Sardinia",
                    "Italy",
                    "Austria",
                    "Czech Republic",
                    "Hungary",
                    "Sweden",
                    "Finland",
                    "Estonia",
                    "Crete",
                    "Balkans",
                    "Turkey",
                    "Russia"))

Table1$N <- table(pops$o_pop)

Table1 <- Table1[order(Table1$Region),c("Population","Region","N","Latitude","Longitude")]
kable(Table1,digits = 3,row.names = F)


```
\pagebreak

##Figure legends

**Figure 1** Sampling locations, heterozygosity

**Figure 2** Population structure in European Great tit populations. **A** Pairwise $F_{ST}$ between all European populations in relation to pairwise geographic distance; **B** pairwise \(F_{ST}\) between a proposed refugial population in Turkey and all other populations, in relation to geographic distance from the refugia. Blue points in **A** and **B** are comparisons involving the island populations of Corsica, Sardinia and Crete, while grey points represent all other comparisons. **C,D** Output from admixture analysis at $K$ = 3, (**C**) and $K$ = 8 (**D**).

**Figure 3** Maximum likelihood tree inferred by Treemix, allowing ten migration events. Migrations events (arrows) are coloured according to their weight (red = higher migration), and horizontal branch lengths are proportional to the amount of genetic drift that has occurred along the branch.  

***Figure 4** 

**Figure 5**

\pagebreak

**Figure 1**

```{R Figure 1,fig.width = 8,fig.height = 4}
Fig1A <- get_map(location = c(-11,31,40,66),zoom = 4,maptype = "watercolor",color = "bw") %>%
  ggmap()+
  geom_point(data = ll,aes(x = Long,y = Lat,size = LD),col = "red")+
  xlab("Longitude")+
  ylab("Latitude")+
  ggtitle("A")

Fig1B <- ggplot(hetlong,aes(x = pop,y = het))+
  geom_boxplot(notch = T,outlier.shape = NA,coef = 0)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90))+
  coord_cartesian(ylim = c(0, 0.6))+
  ylab("Observed heterozygosity")+
  xlab("")+
  ggtitle("B")

multiplot(Fig1A,Fig1B,layout = matrix(c(1:2),nrow = 1))
```







\pagebreak

**Figure 2**

```{r Figure 2,fig.width = 8,fig.height = 10}


Fig2A <- rbind(pd[,c(1:3)],
               data.frame(p1 = unique(c(pd$p1,pd$p2)),p2 = unique(c(pd$p1,pd$p2)),FST = rep(0,length(unique(c(pd$p1,pd$p2))))),
               data.frame(p1 = pd$p2,p2 = pd$p1,FST = pd$FST)) %>%
  mutate(pd,p1 = factor(p1,
       levels = c("Loch_Lomond_Scotland",
                  "Cambridge_UK",
                  "Wytham_UK",
                  "Mariola_Spain",
                  "Font_Roja_Spain",
                  "La_Rouviere_France",
                  "Montpellier_France",
                  "Antwerp_Belgium",
                  "Vlieland_Netherlands",
                  "Westerheide_Netherlands",
                  "Zurich_Switzerland",
                  "Radolfzell_Germany",
                  "Seewisen_Germany",
                  "Pirio_Muro_Corsica",
                  "Sardinia",
                  "Italy",
                  "Vienna_Austria",
                  "Velky_Kosir_Czech_Republic",
                  "Pilis_Mountains_Hungary",
                  "Gotland_Sweden",
                  "Harjavalta_Finland",
                  "Oulu_Finland",
                  "Tartu_Estonia",
                  "Crete",
                  "Bulgaria",
                  "Romania",
                  "Turkey",
                  "Zvenigorod_Russia"))) %>%
    mutate(p2 = factor(p2,
       levels = c("Loch_Lomond_Scotland",
                  "Cambridge_UK",
                  "Wytham_UK",
                  "Mariola_Spain",
                  "Font_Roja_Spain",
                  "La_Rouviere_France",
                  "Montpellier_France",
                  "Antwerp_Belgium",
                  "Vlieland_Netherlands",
                  "Westerheide_Netherlands",
                  "Zurich_Switzerland",
                  "Radolfzell_Germany",
                  "Seewisen_Germany",
                  "Pirio_Muro_Corsica",
                  "Sardinia",
                  "Italy",
                  "Vienna_Austria",
                  "Velky_Kosir_Czech_Republic",
                  "Pilis_Mountains_Hungary",
                  "Gotland_Sweden",
                  "Harjavalta_Finland",
                  "Oulu_Finland",
                  "Tartu_Estonia",
                  "Crete",
                  "Bulgaria",
                  "Romania",
                  "Turkey",
                  "Zvenigorod_Russia"))) %>%
  ggplot(aes(x = p1,y = p2))+
  theme_classic()+
    geom_tile(aes(fill = FST))+
    scale_fill_distiller(palette = "Greens",direction = -1)+
    theme(axis.text.x = element_text(angle = 90))+
  xlab("")+
  ylab("")+
  ggtitle("A")



Fig2B <- ggplot(pd,
                aes(x = Island,y = FST))+
                geom_jitter(col = "grey")+
                theme_bw()+
                ylab(expression(italic(F)[ST]))+
                xlab("")+
  ggtitle("B")


Fig2C <- subset(pd,p1 == "Turkey" | p2 == "Turkey") %>%
          ggplot(aes(x = dist/1000,y = FST))+
            geom_point(col = "grey")+
            theme_bw()+
            theme(legend.position = "none")+
            ylab(expression(italic(F)[ST]))+
            xlab("Distance from Turkey (km)")+
            scale_colour_manual(values = c("darkblue","grey")) +
  annotate("text",
           x = subset(subset(pd,p1 == "Turkey" | p2 == "Turkey"),FST > 0.02)$dist/1000,
           y = subset(subset(pd,p1 == "Turkey" | p2 == "Turkey"),FST > 0.02)$FST-0.003,
           label = revalue(subset(subset(pd,p1 == "Turkey" | p2 == "Turkey"),FST > 0.02)$p1,c(Pirio_Muro_Corsica = "Corsica")),
           size = 3)+
  ggtitle("C")

Fig2D <- structureplot(admix8,pops,8)+
  ggtitle("D")



multiplot(Fig2A,Fig2B,Fig2C,Fig2D,layout = matrix(c(rep(1,10),rep(c(2,3),3),rep(4,6)),nrow = 11,byrow = T))

```


**Figure 3**

```{r Figure 3,fig.width = 6,fig.height = 6}
plot_tree("Data/TreeMix/HapMap_2",mbar = F,cex = 0.8)
```


**Figure 4**
```{r Figure 4,fig.width = 8,fig.height = 7}

fst_admix$ZFST <- (fst_admix$MEAN_FST - mean(fst_admix$MEAN_FST))/sd(fst_admix$MEAN_FST)
fst_cen$ZFST <- (fst_cen$MEAN_FST - mean(fst_cen$MEAN_FST))/sd(fst_cen$MEAN_FST)

fst_both <- rbind(fst_admix,fst_cen)
fst_both$data <- c(rep("Marginal populations",nrow(fst_admix)),rep("Central populations",nrow(fst_cen)))
fst_both$MEAN_FST[fst_both$MEAN_FST < 0] <- 0


Fig4A <- ggplot(fst_both,aes(x = BIN_START,y = ZFST))+
  geom_point(col = "grey",size = 1)+
  theme_classic()+
  geom_smooth(se = F)+
  theme(
        strip.background = element_blank(),
        strip.placement = "outside",
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank())+
  xlab("")+
  ylab(expression(italic(F)[ST])) + 
  facet_grid(data ~ CHROM,scales = "free",space = "free_x",switch = "x") +
  theme(axis.line.y = element_line(),
        axis.line.x = element_line(),
        strip.text.x = element_text(size = 2))+
  ggtitle("A")


recomb$zfst_admix <-  (recomb$fst_admix - mean(recomb$fst_admix))/sd(recomb$fst_admix)
recomb$zfst_cen <-  (recomb$fst_cen - mean(recomb$fst_cen))/sd(recomb$fst_cen)

Fig4B <- ggplot(subset(recomb,MEAN_cM < 100),aes(x = (zfst_admix),y = (zfst_cen)))+
  geom_point(col = "grey")+
  geom_smooth(method = "lm")+
  theme_classic()+
  xlab(expression(italic(F)[`ST (marginal populations)`]))+
  ylab(expression(italic(F)[`ST (central populations)`]))+
  ggtitle("B")

Fig4C <- subset(subset(recomb,MEAN_cM < 100),MEAN_cM > -1) %>%
  ggplot(aes(y = (zfst_admix),x = (MEAN_cM)))+
  geom_point(col = "grey")+
  geom_smooth()+
  theme_classic()+
  ylab(expression(italic(F)[`ST (marginal populations)`]))+
  xlab("log recomb. rate (cM/mb)")+
  ggtitle("C")

Fig4D <- subset(subset(recomb,MEAN_cM < 100),MEAN_cM > -1) %>%
  ggplot(aes(y = (zfst_cen),x = (MEAN_cM)))+
  geom_point(col = "grey")+
  geom_smooth()+
  theme_classic()+
  ylab(expression(italic(F)[`ST (central populations)`]))+
  xlab("log recomb. rate (cM/mb)")+
  ggtitle("D")

multiplot(Fig4A,Fig4B,Fig4C,Fig4D,layout = matrix(c(1,1,1,2,3,4),nrow = 2,byrow = T))


```

**Figure 5**
```{r Figure 5,fig.width = 8,fig.height = 8}
pd <- subset(pd,outlier_slope > -2)

Fig5A <- rbind(pd[,c("p1","p2","outlier_slope")],
               data.frame(p1 = unique(c(pd$p1,pd$p2)),p2 = unique(c(pd$p1,pd$p2)),outlier_slope = rep(0,length(unique(c(pd$p1,pd$p2))))),
               data.frame(p1 = pd$p2,p2 = pd$p1,outlier_slope = pd$outlier_slope)) %>%
  mutate(pd,p1 = factor(p1,
       levels = c("Loch_Lomond_Scotland",
                  "Cambridge_UK",
                  "Wytham_UK",
                  "Mariola_Spain",
                  "Font_Roja_Spain",
                  "La_Rouviere_France",
                  "Montpellier_France",
                  "Antwerp_Belgium",
                  "Vlieland_Netherlands",
                  "Westerheide_Netherlands",
                  "Zurich_Switzerland",
                  "Radolfzell_Germany",
                  "Seewisen_Germany",
                  "Pirio_Muro_Corsica",
                  "Sardinia",
                  "Italy",
                  "Vienna_Austria",
                  "Velky_Kosir_Czech_Republic",
                  "Pilis_Mountains_Hungary",
                  "Gotland_Sweden",
                  "Harjavalta_Finland",
                  "Oulu_Finland",
                  "Tartu_Estonia",
                  "Crete",
                  "Bulgaria",
                  "Romania",
                  "Turkey",
                  "Zvenigorod_Russia"))) %>%
    mutate(p2 = factor(p2,
       levels = c("Loch_Lomond_Scotland",
                  "Cambridge_UK",
                  "Wytham_UK",
                  "Mariola_Spain",
                  "Font_Roja_Spain",
                  "La_Rouviere_France",
                  "Montpellier_France",
                  "Antwerp_Belgium",
                  "Vlieland_Netherlands",
                  "Westerheide_Netherlands",
                  "Zurich_Switzerland",
                  "Radolfzell_Germany",
                  "Seewisen_Germany",
                  "Pirio_Muro_Corsica",
                  "Sardinia",
                  "Italy",
                  "Vienna_Austria",
                  "Velky_Kosir_Czech_Republic",
                  "Pilis_Mountains_Hungary",
                  "Gotland_Sweden",
                  "Harjavalta_Finland",
                  "Oulu_Finland",
                  "Tartu_Estonia",
                  "Crete",
                  "Bulgaria",
                  "Romania",
                  "Turkey",
                  "Zvenigorod_Russia"))) %>%
  ggplot(aes(x = p1,y = p2))+
    geom_tile(aes(fill = outlier_slope))+
    scale_fill_distiller(palette = "Greens",direction = -1)+
    theme(axis.text.x = element_text(angle = 90))+
  xlab("")+
  ylab("")+
  ggtitle("A")



Fig5B <- ggplot(pd,aes(x = dist,y = outlier_slope,col = Island))+
            geom_point()+
  geom_smooth(method = "lm")+
            theme_bw()+
            theme(legend.position = "none")+
            ylab(expression(italic("F")[ST]*" vs recombination coefficient"))+
            xlab("Distance (km)")+
            scale_colour_manual(values = c("darkblue","grey"))+
  ggtitle("B")


Fig5C <- ggplot(pd,aes(x = FST,y = outlier_slope,col = Island))+
            geom_point()+
  geom_smooth(method = "lm")+
            theme_bw()+
            theme(legend.position = "none")+
            ylab(expression(italic("F")[ST]*" vs recombination coefficient"))+
            xlab(expression("Genome-wide "*italic("F")[ST]))+
            scale_colour_manual(values = c("darkblue","grey"))+
  ggtitle("C")



multiplot(Fig5A,Fig5B,Fig5C,layout = matrix(c(rep(1,10),rep(c(2,3),3)),nrow = 8,byrow = T))
```



